No Massive Companion to the Low-Frequency, Coherent Radio-Emitting M~Dwarf GJ\,1151

### 2019-08-29 oldpaper/ms.tex
now as submodule

Rather than explicitly searching for radio emission from known exoplanet hosts, \citet{Callingham_2019} cross-matched sources identified by the LOFAR Two-meter Sky Survey \citep[LoTSS:][]{lotss} with nearby stellar sources found by \gaia, finding the great majority of matches to be chance associations. But by restricting the survey to variable and circularly-polarized emission, the rate of chance associations with background radio galaxies is dramatically reduced. Based on this restricted cross-match, Vedantham et al. (submitted) have detected the quiescent M4.5 dwarf GJ\,1151 at low radio frequencies with LOFAR that has properties that suggest the low-frequency radio emission is driven by a star-planet magnetic interaction. While M~dwarfs are known to flare at low frequencies \citep[e.g.][]{Villadsen19}, this emission lasts over 8\,h and is $64\pm6\%$ circularly polarised. Such emission can be generated by the electron cyclotron maser instability (ECMI) through the interaction of the star with a short ($\sim 1-5$\,day) period planet or a close stellar companion as seen in UV Ceti-like stars \citep{2017ApJ...836L..30L}.

@@

\citet{2019arXiv190505900H} using \kepler estimate that M3-M5 dwarfs host $1.19^{+0.70}_{−0.49}$ planets per star with radii from $0.5-2.5 R_\oplus$ and periods from  0.5-10\,d, the detection of a planet is actually less important than the exclusion of \emph{non}-planetary models, such as emission from a stellar binary interaction. We can confidently say that there is no short-period stellar binary companion to GJ~1151, and that the posterior for the RV trend of $0.02 \pm 0.03$ m/$\text{s}^{-1}\text{d}^{-1}$ is consistent with zero at $1\,\sigma$.

@@

% (1) The star-planet interaction calculations provide a heauristic lower limit on the range of planetary radius (r) and period (p) that are concistent with the properties of the observed radio emission.

% (2) Heuristically speaking, you can trade off planetary radius (r) with orbital distance (d) to get the same radio flux-density. The B-field goes down as d^-2 (parker spiral) and the planetary cross section for the interaction goes as r^2. So the radio data are providing a lower limit on the ratio r/d.

% (3) Now the mass (m) goes as r^3 and period (p) goes as d^2/3, we have a lower limit on m/p^2 from the radio data.

% (4) Again heuristically speaking taking the whole range of possibilities for the precise nature of interaction and the various efficiencies, we can say that an Earth-like planet in a 1-5d period will do the job.

@@

Vedantham et al. (submitted) derive an approximate mass and period estimate for their planet candidate based on the scaling relations of the SPMI. A planet with a larger radius $r$ has a larger cross-section for wind interaction $\propto r^2$, while the B-field drops with semi-major axis $a$ as $\approx a^{-2}$, so that the radio flux only provides a lower limit on $r/d$. Given the mass scaling $\propto r^3$ and orbital period $\propto a^{\frac{2}{3}}$, the radio detection provides a lower limit on $m/p^2$. An Earth-mass planet in a $1-5$\,d orbit is consistent with the observed radio flux, and at a sufficiently short period even a planet with a mass around that of $\sim$Mercury would be sufficient. While the data presented in this Letter conclusively rule out any more massive companion, there is a substantial region of parameter space for these planets that cannot be excluded and the star-planet interaction hypothesis remains reasonable.

>>>>

Rather than explicitly searching for radio emission from known exoplanet hosts, \citet{Callingham_2019} cross-matched sources identified by the LOFAR Two-meter Sky Survey \citep[LoTSS:][]{lotss} with nearby stellar sources found by \gaia, finding the great majority of matches to be chance associations. But by restricting the survey to circularly-polarized emission, the rate of chance associations with background radio galaxies is dramatically reduced. Based on this restricted cross-match, Vedantham et al. (submitted) have detected the quiescent M4.5 dwarf GJ\,1151 at low radio frequencies with LOFAR that has properties that suggest the low-frequency radio emission is driven by a star-planet magnetic interaction. While M~dwarfs are known to flare at low frequencies \citep[e.g.][]{Villadsen19}, this emission lasts over 8\,h and is $64\pm6\%$ circularly polarised. Such emission can be generated by the electron cyclotron maser instability (ECMI) through the interaction of the star with a short ($\sim 1-5$\,day) period planet or a close stellar companion as seen in UV Ceti-like stars \citep{2017ApJ...836L..30L}.

@@

A non-detection of a planet here does not rule out a planetary origin for the emission, as planets are common around such stars: \citet{2019arXiv190505900H} using \kepler estimate that M3-M5 dwarfs host $1.19^{+0.70}_{−0.49}$ planets per star with radii from $0.5-2.5 R_\oplus$ and periods from  0.5-10\,d. 
The detection of a planet is actually less important than the exclusion of \emph{non}-planetary models, such as emission from a stellar binary interaction. We can confidently say that there is no short-period stellar binary companion to GJ~1151, and that the posterior for the RV trend of $0.02 \pm 0.03$ m/$\text{s}^{-1}\text{d}^{-1}$ is consistent with zero at $1\,\sigma$.

@@

@@

Vedantham et al. (submitted) derive an approximate mass and period estimate for their planet candidate based on the scaling relations of the SPMI. A planet with a larger radius $r$ has a larger cross-section for wind interaction $\propto r^2$, while the stellar Poynting flux at the location of the planet drops with semi-major axis $a$ as $\approx a^{-2}$, so that the radio flux provides a lower limit on $r/d$. Given the planetary mass scaling $\propto r^3$ and orbital period $\propto a^{\frac{2}{3}}$, the radio detection provides a lower limit on $m/p^2$.

Considering a plausible range of interaction and emission efficiencies, Vedantham et al. estimate that an Earth-mass planet in a $\sim 1-5$ day orbit can generate the observed radio flux. At sufficiently large efficiencies, even a Mercury-mass planet also remains plausible. While the data presented in this Letter conclusively rule out any more massive companion, there is a substantial region of parameter space for these planets that cannot be excluded and the star-planet interaction hypothesis remains reasonable.
